### 2018-07-09 latex/introduction.tex
Inserted correct references and did some fixing

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. Even though best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}. These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio. Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.

The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated. It has now been established, however, that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018}. This  suggests that NAD-dependent signalling reactions consume substantial amounts of NAD. Therefore, we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD homeostasis have been linked to various mainly age related diseases, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}. Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway staring with the convertion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{PMID:19666527}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. Although it has been suggested that the process potentially acts as a metabolic methylation sink \cite{PMID:28291578} .

Secondly, in many prokayotes as well as in plants and fungi a four steps starting with the deamidation of Nam to nicotinic acid (NA) by NADA is used. (\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT. The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{PMID:21729004}. The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This reaction includes an enzyme adenylation step that consumes ATP. Therefore the Nam recycling by NADA is energetically less efficient under normal conditions.

We here present a phylogenetic analysis of the NAD pathways. Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost. This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam. Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT. Simulating the resource competition, we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.

>>>>

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}. These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio. Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.

The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated. It has now been established, however, that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018}. This suggests that NAD-dependent signalling reactions consume substantial amounts of NAD. Therefore, we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell-type-specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD homeostasis have been linked to various mainly age related diseases, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}. Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.

Secondly, in many prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used. (\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT. The convertion of the mononucleotides NMN and NAMN into their dinucleotides NAD and NAAD is performed by the same enzyme, the Nam/NA adenylytransferases (NMNATs) that is essential in all organisms \cite{DeFigueiredo2011}. The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This final reaction includes an enzyme adenylation step that consumes ATP. Therefore the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.

We here present a phylogenetic analysis of the NAD pathways. Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, is lost. This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam. Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT. Simulating the resource competition, we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.

### 2018-07-09 latex/results.tex
Inserted correct references and did some fixing

\subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
We have  previously shown  that NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor \cite{PMID:21729004}. In addition, the Preiss-Handler pathway was found  present in virtually all organisms included in an earlier phylogenetic analysis\cite{Gossmann2012FEBS}. Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways, NADA and NamPRT. These show a scattered distribution in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur concomitantly in some marine invertebrates \cite{Gossmann2012FEBS}. In addition, we included  NAM methyltransferase NNMT into our analysis, as its function and origin is unclear. 
As can be seen from Fig. (fig.~\ref{fig:phylo_distribution}A, bacteria, fungi and plants predominantly encode NADA and only a few of them harbour NamPRT.  In contrast, Metazoa predominantly have NamPRT together with NNMT. Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT dominating and NADA appears to be lost in vertebrates, in most organisms NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. As we were unable to find any \textit{NNMT} gene with an e-value below .... \todo{Mathias B. include e-value}  in fungi or plants. The gene with the highest sequence similarity in S. cerevisiae is  ..... with proven function .....\todo{Mathias B: include e-value}

Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT. In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds. The reason why about half of the sequenced bird genomes do not seem to encode for  NNMT,  remains unclear. The distribution of NNMT in birds is quite scattered (suppl. fig.~S2). not corresponding to current species trees \cite{http://dx.doi.org/10.1038/nature15697} . The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine. There few Metazoan species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1. The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade. With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in Protostomes, whereas most Deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway. Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling. This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis. However, NNMT irreversibly removes Nam from salvage and thus is assumed to potentially decrease the cellular NAD-concentration.

@@

To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes, we initially used equal amounts for all enzymes. As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes.  Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites. As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.

First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families. We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT. To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use 10 fold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme.  As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B). The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).

These results can be explained by looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1. Most NAD-consuming enzymes are inhibited by their product Nam. Thus the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is of course lower than in the system without NNMT.

@@

Figure~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT. At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E). If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it. The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}F).

As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate. In figure~\ref{fig:optimal_substr_affinities} is shown that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any further  advantage.

@@

As our simulations suggested that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in NamPRT upon the appearance of NNMT. The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1. We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed. The sequence corresponds to a loop at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g. structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}). The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{PMID:19666527}.

From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function. The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS. To test this hypothesis, we recombinatly expressed FLAG-tagged  wildtype and mutant NamPRT in Hela S3 cells. Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C). Thus, the presence of the loop does not compromise the nuclear localisation.

The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations. We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}. After purification (suppl. fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP. The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity. In contrast to wildtype NamPRT, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).

>>>>

\subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}

We have previously shown that NMNAT is essential for NAD biosynthesis independent of the biosynthetic precursor or pathway \cite{DeFigueiredo2011}. In addition, the Preiss-Handler pathway was found  present in virtually all organisms included in an earlier phylogenetic analysis \cite{Gossmann2012FEBS}. Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT. The phylogenetic distribution of the two enzymes is scattered in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}. In addition, we included the Nam methyltransferase NNMT into our analysis, as its function and origin is unclear.

As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT. In contrast, Metazoa predominantly lost NADA amd have NamPRT together with NNMT. NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B). As we were unable to find any \textit{NNMT} gene with an e-value below xxx \todo{Mathias B. include e-value} in fungi or plants. The gene with the highest sequence similarity in \textit{S.~cerevisiae} is ... with proven function ... \todo{Mathias B: include values}.

Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT. In deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds. The reason why about half of the sequenced bird genomes do not seem to encode for NNMT remains unclear. The distribution of NNMT in birds is quite scattered (suppl. fig.~S2). \todo{figure not corresponding to current species trees \cite{Prum2015}} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine. There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1. The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade. With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway. Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling. This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.

@@

To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we had limited information about expression levels of enzymes, we initially used equal amounts for all enzymes. As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes. Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites. As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.

First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families. We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT. To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use tenfold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme. As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B). The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).

These results can be explained by looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1. Most NAD-consuming enzymes are inhibited by their product Nam. Thus the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.

@@

Figures~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT. At an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E). If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it. The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}F).

As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate. The substrate affinity values found in the human enzymes (indicated by black asterisks in fig.~\ref{fig:optimal_substr_affinities}) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any further advantage.

@@

As our simulations suggested that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in NamPRT upon the appearance of NNMT. The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1. We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed. The sequence corresponds to a loop at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g. structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}). The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009}.

From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function. The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS. To test this hypothesis, we created a mutant NamPRT lacking the loop and recombinatly expressed FLAG-tagged wildtype and mutant NamPRT in Hela S3 cells. Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C). Thus, the presence of the loop does not compromise the nuclear localisation.

The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations. We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}. After purification (suppl. fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP. The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity. In contrast to wildtype NamPRT, the mutant did not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
